Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2015-09-21 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
correct streamfunction signs again, describe 4th order difference, TV analysis for nonorthogonality

\section{Model}
\label{sec:model}
Three models are used for the test cases in this study, and all are operated in a two-dimensional $x$--$z$ slice configuration.  The first model discretises the linear advection equation in flux form:
\begin{align}
\partial \phi / \partial t + \nabla \cdot \left( \mathbf{u} \phi \right) = 0
\label{eq:advect}
\end{align}
where the tracer density, $\phi$, is interpolated onto cell faces using one of two schemes: first, the centred linear scheme which takes the average of the two neighbouring cell values; second, the upwind-biased multidimensional cubic scheme from \citet{weller-shahrokhi2014} which is non-monotonic and not flux corrected.
The time derivative is solved using a three-stage, second order Runge-Kutta scheme defined as:
\begin{subequations}
\begin{align}
	\phi^\star &= \phi^{(n)} + \Delta t f(\phi^{(n)}) \\
	\phi^{\star\star} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^\star) \right) \\
	\phi^{(n+1)} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^{\star\star}) \right)
\end{align}
\end{subequations}
where \(f(\phi^{(n)}) = - \nabla \cdot (\mathbf{u} \phi^{(n)})\) at time level \(n\). 
To ensure that the discrete velocity field is non-divergent, velocities are prescribed at cell faces by differencing the streamfunction, \(\Psi(x, z)\), along the edges from \(\Psi\) stored at cell vertices.

The second model is a modified version of the linear advection finite difference model first used by \citet{schaer2002}.  It uses terrain following coordinates and it is configured with leapfrog timestepping and a fourth-order centred difference scheme: \TODO{this wasn't properly documented in schaer2002 but we can document it using the Fortran code}
\begin{align}
	TODO
\end{align}

@@

The third model is taken from \citet{weller-shahrokhi2014} which details a discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \mathbf{u}}{\partial t} + \nabla \cdot \rho \mathbf{u}\otimes\mathbf{u} &= \rho \mathbf{g} - c_p \rho \theta \nabla \Pi - \mu \rho \mathbf{u} \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \mathbf{u} &= 0 \label{eq:cont} \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \mathbf{u} \theta &= 0 \label{eq:theta} \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0} \label{eq:state}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\mathbf{u}\) is the velocity field, \(\mathbf{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the temperature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.  \(\mu\) is a dimensionless damping function used for the sponge layer in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.

@@

A series of two-dimensional tests are performed over idealised orography.  For each test, results on the BTF, SLEVE and cut cell grid are compared.  The first test from \citet{schaer2002} advects a tracer in a horizontal velocity field.  Second, a new tracer advection test is formulated employing a terrain following velocity field to challenge the advection scheme on cut cell grids.  The third test solves the Euler equations for a stably stratified atmosphere initially at rest, following \citet{klemp2011}.  Finally, as specified by \citet{schaer2002}, a test of orographically-induced gravity waves is performed.  No explicit diffusion is used in any of the tests.

@@

\TODO{Schaer defines u=-dPsi/dz, w=dPsi/dx --- check all this streamfunction stuff AGAIN}

The discrete velocity field is defined using a streamfunction, \(\Psi\).  Since \(u = \partial \Psi / \partial z\), the streamfunction is found by vertical integration of the velocity profile:
\begin{align}
	\Psi(z) &= \frac{u_0}{2} \left\{ \begin{array}{l l}
		\left( 2z - z_1 - z_2 \right) & \enskip \text{if $z > z_2$} \\
		z - z_1 - \frac{z_2 - z_1}{\pi} \sin \left(\pi \frac{z - z_1}{z_2-z_1}\right) & \enskip \text{if $z_1 < z \leq z_2$} \\
		0 & \enskip \text{if $z \leq z_1$}
	\end{array} \right.
\end{align}

@@

The centred linear scheme is sensitive to changes in the terrain profile: when the domain width is reduced to \SI{300}{\kilo\meter}, the $\ell_2$ error on the BTF grid rises from \num{0.284} to \num{0.432}.  The upwind-biased cubic scheme is less sensitive to the same change, with the $\ell_2$ error decreasing from \num{0.107} to \num{0.104}.

@@

The spatial domain, mountain profile, initial tracer profile and discretisation are the same as those in the horizontal tracer advection test.  The velocity field is defined using a streamfunction, $\Psi$, so that the continuous velocity field is non-divergent and follows the BTF coordinate surfaces given by equation~\ref{eqn:btf} such that
\begin{equation}
	\Psi(x,z) = u_0 H \frac{z - h}{H - h} \label{eqn:streamfunc-btf}
\end{equation}
where $u_0 = \SI{10}{\meter\per\second}$, which is the horizontal wind speed where $h(x) = 0$.
The horizontal and vertical components of velocity, $u$ and $w$, are then given by
\begin{align}
	u &= \frac{\partial \Psi}{\partial z} = u_0 \frac{H}{H - h}, \quad w = -\frac{\partial \Psi}{\partial x} = u_0 H \frac{\mathrm{d} h}{\mathrm{d} x} \frac{H - z}{\left( H - h \right)^2} \label{eqn:uw-btf} \\
	\frac{\mathrm{d} h}{\mathrm{d} x} &= - h_0 \left[ 
		\beta \cos^2 \left( \alpha x \right) \sin \left( 2 \beta x \right) +
		\alpha \cos^2 \left( \beta x \right) \sin \left( 2 \alpha x \right)
	\right]
\end{align}
Unlike the horizontal advection test, flow extends from the top of the domain all the way to the ground.  The discrete velocity field is calculated using the streamfunction in the same way as the horizontal advection test.

@@

\subsection{Terrain following advection of thermal profile}
The spatial domain, mountain profile, grid resolutions and timesteps are the same as those in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.  The terrain following velocity field is defined by the streamfunction:
\begin{align}
	\Psi(x,z) = u_0 \left\{ \begin{array}{l l}
			H_\mathrm{TF} \frac{z - h}{H_\mathrm{TF} - h} & \enskip \text{if $z \leq H_\mathrm{TF}$} \\
			z & \enskip \text{if $z > H_\mathrm{TF}$}
	\end{array} \right.
\end{align}
where \(H_\mathrm{TF} = \SI{20}{\kilo\meter}\) is the level at which the terrain following layers become flat; the domain height is \(\SI{30}{\kilo\meter}\).
For $z \leq H$, the $u$ and $w$ components of velocity are given by equation~\ref{eqn:uw-btf}, but $h(x)$ has the same form as equation~\ref{eqn:resting:mountain}, hence the derivative is:
\begin{align}
	\frac{\mathrm{d} h}{\mathrm{d} x} = -h_0 \exp \left(- \left( \frac{x}{a} \right)^2 \right) \left[ \frac{\pi}{\lambda} \sin \left( \frac{2 \pi x}{\lambda} \right) - \frac{2x}{a^2} \cos^2\left( \frac{\pi x}{\lambda} \right) \right]
\end{align}
For $z > H$, $u = u_0$ and $w = 0$.

>>>>

\section{Models}
\label{sec:model}
Three models are used for the test cases in this study: two linear advection models and a model of the fully-compressible Euler equations.  All are operated in a two-dimensional $x$--$z$ slice configuration.

\subsection{Finite volume linear advection model}
The first model discretises the linear advection equation in flux form:
\begin{align}
\partial \phi / \partial t + \nabla \cdot \left( \mathbf{u} \phi \right) = 0
\label{eq:advect}
\end{align}
where $\mathbf{u} = (u, w)$ is a prescribed velocity field and the tracer density, $\phi$, is interpolated onto cell faces using one of two schemes: first, the centred linear scheme which takes the average of the two neighbouring cell values; second, the upwind-biased multidimensional cubic scheme from \citet{weller-shahrokhi2014} which is non-monotonic and not flux corrected.
The time derivative is solved using a three-stage, second order Runge-Kutta scheme defined as:
\begin{subequations}
\begin{align}
	\phi^\star &= \phi^{(n)} + \Delta t f(\phi^{(n)}) \\
	\phi^{\star\star} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^\star) \right) \\
	\phi^{(n+1)} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^{\star\star}) \right)
\end{align}
\end{subequations}
where \(f(\phi^{(n)}) = - \nabla \cdot (\mathbf{u} \phi^{(n)})\) at time level \(n\). 
To ensure that the discrete velocity field is non-divergent, velocities are prescribed at cell faces by differencing the streamfunction, \(\Psi(x, z)\), along the edges from \(\Psi\) stored at cell vertices.

\subsection{Finite difference linear advection model}
The second model is a modified version of the linear advection model first used by \citet{schaer2002}.  It uses terrain following coordinates and it is configured with leapfrog timestepping and a fourth-order centred difference scheme:
\begin{align}
	\frac{\partial \phi_i}{\partial x} \approx \frac{1}{\Delta x} \left( \frac{1}{12} \phi_{i-2} - \frac{2}{3} \phi_{i-1} + \frac{2}{3} \phi_{i+1} - \frac{1}{12} \phi_{i+2} \right)
\end{align}

@@

\subsection{Fully-compressible model}
The third model is taken from \citet{weller-shahrokhi2014} which details a discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \mathbf{u}}{\partial t} + \nabla \cdot \rho \mathbf{u}\otimes\mathbf{u} &= \rho \mathbf{g} - c_p \rho \theta \nabla \Pi - \mu \rho \mathbf{u} \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \mathbf{u} &= 0 \label{eq:cont} \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \mathbf{u} \theta &= 0 \label{eq:theta} \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0} \label{eq:state}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\mathbf{u}\) is the velocity field, \(\mathbf{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the temperature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.  \(\mu\) is a dimensionless damping function used for the sponge layer in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.

@@

A series of two-dimensional tests are performed over idealised orography.  For each test, results on the BTF, SLEVE and cut cell grid are compared.  The first test from \citet{schaer2002} advects a tracer in a horizontal velocity field.  Second, a new tracer advection test is formulated employing a terrain following velocity field to challenge the advection scheme on cut cell grids.  The third test solves the Euler equations for a stably stratified atmosphere initially at rest, following \citet{klemp2011}.  Finally, as specified by \citet{schaer2002}, a test of orographically-induced gravity waves is performed.  \TODO{advection of thermal profile} No explicit diffusion is used in any of the tests.

@@

The discrete velocity field is defined using a streamfunction, \(\Psi\).  Given that \(u = -\partial \Psi / \partial z\), the streamfunction is found by vertical integration of the velocity profile:
\begin{align}
	\Psi(z) &= -\frac{u_0}{2} \left\{ \begin{array}{l l}
		\left( 2z - z_1 - z_2 \right) & \enskip \text{if $z > z_2$} \\
		z - z_1 - \frac{z_2 - z_1}{\pi} \sin \left(\pi \frac{z - z_1}{z_2-z_1}\right) & \enskip \text{if $z_1 < z \leq z_2$} \\
		0 & \enskip \text{if $z \leq z_1$}
	\end{array} \right.
\end{align}

@@

The centred linear scheme is sensitive to changes in the terrain profile: when the domain width is reduced to \SI{300}{\kilo\meter}, the $\ell_2$ error on the BTF grid rises from \num{0.284} to \num{0.432}.  The same sensitivity is found using terrain following coordinates and centred differences in space and time, as implemented by \citet{schaer2002}.  This sensitivity can be explained by realising that increased non-orthogonality or flow misalignment tends to increase errors.  To measure non-orthogonality, consider the total variation, $TV$, of the discrete terrain profile:
\begin{align}
	TV = \sum_i | h(x_{i+1}) - h(x_i) |
\end{align}
where $x_i$ is the horizontal location of a vertex on the lower boundary.  For a domain width of \SI{301}{\kilo\meter}, $TV = \SI{18095}{\meter}$, but for a width of \SI{300}{\kilo\meter}, $TV = \SI{18964}{\meter}$.
The upwind-biased cubic scheme is less sensitive to the same change, with the $\ell_2$ error decreasing from \num{0.107} to \num{0.104}.

@@

The spatial domain, mountain profile, initial tracer profile and discretisation are the same as those in the horizontal tracer advection test.  The velocity field is defined using a streamfunction, $\Psi$, so that the continuous velocity field is non-divergent and follows the BTF coordinate surfaces given by equation~\ref{eqn:btf} such that
\begin{equation}
	\Psi(x,z) = -u_0 H \frac{z - h}{H - h} \label{eqn:streamfunc-btf}
\end{equation}
where $u_0 = \SI{10}{\meter\per\second}$, which is the horizontal wind speed where $h(x) = 0$.
The horizontal and vertical components of velocity, $u$ and $w$, are then given by
\begin{align}
	u &= -\frac{\partial \Psi}{\partial z} = u_0 \frac{H}{H - h}, \quad w = \frac{\partial \Psi}{\partial x} = u_0 H \frac{\mathrm{d} h}{\mathrm{d} x} \frac{H - z}{\left( H - h \right)^2} \label{eqn:uw-btf} \\
	\frac{\mathrm{d} h}{\mathrm{d} x} &= - h_0 \left[ 
		\beta \cos^2 \left( \alpha x \right) \sin \left( 2 \beta x \right) +
		\alpha \cos^2 \left( \beta x \right) \sin \left( 2 \alpha x \right)
	\right]
\end{align}
Unlike the horizontal advection test, flow extends from the top of the domain all the way to the ground.  The discrete velocity field is calculated using the streamfunction in the same way as the horizontal advection test.

@@

\subsection{Terrain following advection of thermal profile}
The spatial domain, mountain profile, grid resolutions and timesteps are the same as those in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.  The terrain following velocity field is defined by the streamfunction:
\begin{align}
	\Psi(x,z) = -u_0 \left\{ \begin{array}{l l}
			H_\mathrm{TF} \frac{z - h}{H_\mathrm{TF} - h} & \enskip \text{if $z \leq H_\mathrm{TF}$} \\
			z & \enskip \text{if $z > H_\mathrm{TF}$}
	\end{array} \right.
\end{align}
where \(H_\mathrm{TF} = \SI{20}{\kilo\meter}\) is the level at which the terrain following layers become flat; the domain height is \(\SI{30}{\kilo\meter}\).
For $z \leq H$, the $u$ and $w$ components of velocity are given by equation~\ref{eqn:uw-btf}, but $h(x)$ has the same form as equation~\ref{eqn:resting:mountain}, hence the derivative is:
\begin{align}
	\frac{\mathrm{d} h}{\mathrm{d} x} = -h_0 \exp \left(- \left( \frac{x}{a} \right)^2 \right) \left[ \frac{\pi}{\lambda} \sin \left( \frac{2 \pi x}{\lambda} \right) - \frac{2x}{a^2} \cos^2\left( \frac{\pi x}{\lambda} \right) \right]
\end{align}
For $z > H$, $u = u_0$ and $w = 0$.
